### 2012-11-15 paper/isca40format/Msi.tex
stuff

\section{Informal description of the proof}
\label{sec:informal-proof}

\newcommand{\msg}[5]{\text{$#1 \langle #2 \rightarrow #3, #4, #5 \rangle$}}
\newcommand{\Req}[4]{\msg{Req}{#1}{#2}{#3}{#4}}
\newcommand{\Resp}[4]{\msg{Resp}{#1}{#2}{#3}{#4}}
\newcommand{\Data}[4]{\msg{Data}{#1}{#2}{#3}{#4}}
\newcommand{\Mesg}[4]{\msg{Mesg}{#1}{#2}{#3}{#4}}
\newcommand{\FirstInv}{\emph{FirstInv}}
%\input{Atomic}

Let us walk through our design methodology for our message-passing hierarchical
directory based protocols.

\subsection{Message-passing directory-based MSI protocol for 2-levels, no
replacement evictions}
For the sake of explanation, let us start with an MSI protocol, with just two
levels of the memory hierarchy -- the L1 caches and the main memory. We will
eventually build realistic protocols with various optimizations.

Let us assume that the caches have enough capacity so that no line gets evicted
for the sake of replacement. Lines may however get evicted to enforce the
single-writer property for cache coherence, \ie when one L1 cache has
permission to modify the data for an address, no other cache has read or write
permission for that address.

An L1 cache stores the information given in Figure \ref{table:lineinfo}. The
coherence state for an address denotes the permission that a cache has for the
address: $M = \{R, W\}, S = \{R\}, I = \{\}$. State $I$ is akin to the cache not
having the data corresponding to $a$, since it can neither read or write that
cache line.

\begin{figure}
\begin{tabularx}{\linewidth}{|cX|}
\hline
$state[c][a]$ & coherence state (from set $\{M, S, I\}$) of cache $c$ for
address $a$\\
$data[c][a]$ & data in cache $c$ for address $a$\\
\hline
\end{tabularx}
\caption{Values stored by a cache for an address}
\label{table:lineinfo}
\end{figure}

In order to serve a request from the processor, if a cache has to \emph{upgrade}
its state for address $a$, \ie change from $x$ to $y$ where $x > y$
(Figure \ref{table:fns}), it can do so only after ensuring that the new state is
\emph{compatible} with the states of other caches for that address, \ie other
caches are in a state less than or equal to $toCompatible(y)$
(Figure \ref{table:fns}). This is needed to ensure the single-writer property of
the MSI protocol -- if a cache's state is $M$ for an address $a$, every other cache
must be in $I$ for address $a$. But, the caveat is that the L1 cache can access
only its own state and data, as opposed to that of other caches.

\begin{figure}
\begin{tabularx}{\linewidth}{|cX|}
\hline
$toCompatible(s)$ & If the state of one cache is $s$ for address $a$, the
highest state of another cache for the same $a$
$\newline
toCompatible(M) = I,\newline
toCompatible(S) = S,\newline
toCompatible(I) = M
$\\
\hline
$x > y$ & Denotes if state $x$ has more permissions than state $y$.
$M > S, M > I, S > I$. Relations $\le, >, \ge$ are also defined appropriately\\
\hline
\end{tabularx}
\caption{Abstraction functions}
\label{table:fns}
\end{figure}

The memory controller has access to the data from the memory but not to the
state and data of the L1 caches. Instead, it only has access to a directory
which has a version of the coherence states of the caches for each address. The
version of the state in the directory may not be up to date, \ie may not be the
same as the one in the caches, all the time. Figure \ref{table:dirinfo} shows
the information stored in the memory and directory. We will use the word memory
and directory interchangeably for the 2-level protocol. The caches can not
communicate with each other but can communicate with the directory via messages.

@@

Let's us assume that we have a system which guarantees the invariants given
in Figure \ref{FirstInv}.
\floatstyle{boxed}
\restylefloat{figure}

\begin{figure}
\begin{enumerate}
\item The cache can upgrade $state[c][a]$ from $x$ to $y$ where $y > x$, \ie
\emph{downgrade} its directory state only after it receives the appropriate
authorization from the directory for address $a$. \label{cwaitresp}
\item If the cache wants to upgrade $state[c][a]$ to $x$, there is a
message \Req{c}{Dir}{a}{x} which cache $c$ can send to the directory requesting
the upgrade authorization, and the cache will eventually be able to send
the request. \label{csendreq}
\item If the directory receives an upgrade request \Req{c}{Dir}{a}{x} from cache
$c$ for address $a$, then just before receiving the request, $dir[c][a] < x$.
\label{drecvreq}
\item The directory can change $dir[c][a]$ from $x$ to $y$ where $y < x$, \ie
\emph{downgrade} its directory state only after it receives the appropriate
authorization from cache $c$ for address $a$. \label{dwaitresp}
\item If the directory wants to downgrade $dir[c][a]$ to $x$, there is a
message \Req{Dir}{c}{a}{x} which the directory can send to cache $c$ requesting
the downgrade authorization, and the directory will eventually be able to send
the request. \label{dsendreq}
\item If the directory has sent a downgrade request for an address $a$ to cache
$c$, it will eventually get the downgrade authorization response for address $a$
from cache $c$ to downgrade to $x$, where $x$ is less than or equal to the state
to which the directory sent a request to downgrade to. \label{drecvresp}
\item If $dir[c][a] = M$ just before receiving a downgrade authorization from
cache $c$ for address $a$, then the cache would have sent the data along with
the authorization. \label{drecvdata}
\item If the directory wants to send an authorization message to upgrade the
state of address $a$ in cache $c$ to $x$ (\Resp{Dir}{c}{a}{x}), it will
eventually be able to send it. \label{dsendresp}
\item The directory can send the data from the memory along with the upgrade
authorization message if it wants to. \label{dsenddata}
\item The directory's version of the cache's state is always conservative, \ie
$\forall c, a,\; dir[c][a] \ge state[c][a]$. \label{conservative}
\end{enumerate}
\caption{Initial set of invariants \FirstInv}
\label{FirstInv}
\end{figure}

\floatstyle{plain}
\restylefloat{figure}

We will walk through how the MSI coherence protocol (with no replacement
evictions) can be implemented using the above invariants and argue why it is a
correct implementation.

The directory's role is to ensure the single-writer property and that the last
updated data is being read. In order to implement the cache coherence protocols
using the invariants of Figure \ref{FirstInv}, the directory's ensures that at
all times, if for some $c, dir[c][a] = M$, the for every other $c', dir[c'][a] =
I$.

If a cache gets a request from the core which can not be handled because of lack
of permissions, it (eventually) sends an upgrade request to the directory.
\FirstInv.\ref{csendreq} ensures that such a request can eventually be sent.

If the directory receives a request from cache $c$ to upgrade the state of
address $a$ to $x$, because of \FirstInv.\ref{drecvreq}, $dir[c][a] < x$ and
since $cache[c][a] \le dir[c][a]$ because of \FirstInv.\ref{conservative},
$cache[c][a] < x$ and hence the directory must send an upgrade authorization
response. The directory sends a request \Req{Dir}{c'}{a}{toCompatible(x)} to
every cache $c'$ for which $dir[c'][a] > toCompatible(x)$. Such a request can be
sent eventually because of \FirstInv.\ref{dsendreq}.

Because of \FirstInv.\ref{drecvresp}, the directory will eventually receive
responses from each of the requested caches $c'$
(\Resp{c'}{Dir}{a}{toCompatible(x)}) authorizing the downgrade of $dir[c'][a]$
to $toCompatible(x)$. If $dir[c'][a] = M$ just before receiving the
authorization, then by \FirstInv.\ref{drecvdata}, the directory would also
receive data from $c'$. Because the directory maintains the single-writer
invariant, every other cache $c''$ would have $dir[c''][a] = I$ just before
receiving the authorization. This implies for every other cache $c''$,
$state[c''][a] = I$ by the \FirstInv.\ref{conservative} invariant, and hence
$c'$ has the most up-to-date copy for address $a$. Thus the memory is updated
with the received up-to-date copy for address $a$.

Once all the authorizations have been received, for every cache $c'$,
$dir[c'][a] \le toCompatible(x)$ (other than the one that made the request $c$).
Since the \FirstInv.\ref{conservative} invariant ensures that $cache[c'][a] \le
dir[c'][a]$, we have $cache[c'][a] \le toCompatible(x)$, ensuring the
single-writer property (because $toCompatible(M) = I$). The directory can now
change $dir[c][a]$ to $x$ and send \Resp{Dir}{c}{a}{x} authorizing $c$ to
upgrade state of address $a$ to $x$ (\FirstInv.\ref{dsendresp} ensures that such
a message can be sent). Just before sending the authorization, if $dir[c][a] =
I$, it implies that $state[c][a] = I$ because of the
\FirstInv.\ref{conservative} invariant.  Thus the cache doesn't have data
corresponding to address $a$ and the directory has to supply the most up to date
version of the address (from memory, since the memory is up to date), which it
can because of \FirstInv.\ref{dsenddata}.

The \FirstInv\ invariants essentially makes it straight forward to design the
behavior of the directory . For the sake of argument, let
\FirstInv.\ref{drecvreq} not be guaranteed. Consider a scenario the cache $c$
which has sent a request \Req{c}{Dir}{a}{M} earlier but hasn't received the
upgrade authorization response yet. Meanwhile, $c$ sends another upgrade request
\Req{c}{Dir}{a}{S}. Let us assume that the directory processed the requests in
the order in which they were sent. Once the directory has finished processing
the first request, it changes its directory state $dir[c][a]$ to $M$. When it
receives the second request, it is not easy to decide if the directory has to
drop the request or to send another response authorizing the cache to change to
$S$.

The difficulty in message passing protocols comes from the fact that it is
difficult to guarantee the above invariants, even for a 2-level memory hierarchy
discussed above as shown in the examples below.

\subsubsection*{Example 1: Inability to send messages because of network
congestion:}
For instance, the ability of a cache or the directory to send messages
(\FirstInv.\ref{dsendreq}, \FirstInv.\ref{dsendresp}, \FirstInv.\ref{csendreq},
\etc) depends upon the availability of free buffers in the network; if the
network remains congested all the time, then it would not be possible to send the
messages, and the system would deadlock.

\subsubsection*{Example 2: Multiple upgrade requests from the cache for the same
address:}
If a cache is allowed to make multiple requests to upgrade the state of an
address (say to $M$ and then to $S$) before receiving the upgrade
authorizations, then it might violate \FirstInv.\ref{drecvreq} if the requests
were processed by the directory in the order they were received.

\subsubsection*{Example 3: On sending an upgrade request, the cache waits till
it gets an authorization:}
If the behavior of the caches were such that if a cache $c$ has sent an
upgrade request \Req{c}{Dir}{a}{x} to the directory, it keeps waiting for the
upgrade authorization response \Resp{Dir}{c}{a}{x} before processing any
requests from the directory for address $a$, then the cache coupled with our
simple directory implementation would deadlock. Consider a scenario with two
caches $X$ and $Y$, with both the caches in state $S$ for address $a$. Both the
caches got a store request from the processor and want to upgrade their states
to $M$, and hence send \Req{X}{Dir}{a}{M} and \Req{Y}{Dir}{a}{M}, respectively,
to the directory. Let us say the directory receives \Req{Y}{Dir}{a}{M} first,
and in order to process it, according to our implementation of the directory,
sends a downgrade request \Req{Dir}{X}{a}{I} to cache $X$. But the cache 
does not process any request from the directory for address $a$, according to
our assumption, and hence will never send the downgrade authorization
\Resp{X}{Dir}{a}{I} to the directory, violating \FirstInv.\ref{drecvresp}, and
hence causing a deadlock.

\subsubsection*{Example 4: At the directory, authorization responses are blocked
by upgrade requests:}
Consider a scenario where two caches $X$ and $Y$ are both in state $S$ for address
$a$. $Y$ got a store request from the processor and sends an upgrade request
\Req{Y}{Dir}{a}{M} to the directory. The directory receives that request, and
sends a downgrade request \Req{X}{Dir}{a}{I} to $X$. $X$ sends a downgrade
authorization \Resp{X}{Dir}{a}{I} to the directory on receiving the downgrade
request. But the directory can not receive the authorization because it is
blocked by other requests (including \Req{Y}{Dir}{a}{M}). This violates
\FirstInv.\ref{drecvresp} and hence creates a deadlock.

These examples show that if the \FirstInv\ invariants are violated, then it
either breaks cache coherence (by creating deadlocks) or it makes correctly
designing the directory very difficult.

Before we show how the \FirstInv\ invariants can be guaranteed (which requires
the description of the behavior of the L1 caches), let us consider a
slightly more complicated system where caches can evict lines for replacement.

\subsection{Message-passing directory-based MSI protocol for 2-levels, with
replacement evictions}
We again have just two levels of memory hierarchy -- the L1 caches and the main
memory. But the caches have finite capacity. If receives a request from the core
for an address $a$ and has no space to accommodate address $a$, it evicts
another address $a'$ to make room for $a$.

In the previous scenario, where the caches didn't have to evict for replacement,
a cache only had to send downgrade authorizations to the directory on receiving
a corresponding request. But in the new scenario, a cache can send a downgrade
notification to the directory on evicting an address. Such a notification is
useful in keeping the directory's version of the cache's state up to date, so we
will restrict the family of directory-based protocols that we discuss to those
where the caches send such a notification, voluntarily.

The directory handles this notification in the exact manner by which it handled
a downgrade authorization response from the cache, so the notification message
is the same as a downgrade authorization message from cache $c$ for authorizing
the downgrade of the $dir[c][a]$ to $I$ (\Resp{c}{Dir}{a}{I}).

We will now discuss a behavior of the L1 cache that would violate the
invariant that the directory's version of the states of the caches for each
address is conservative.

\subsubsection*{Example 5: Cache sends voluntary notifications
(\Resp{c}{Dir}{a}{I}) indiscriminately}
Consider a system with two caches $X$ and $Y$. Let $state[X][a] = S, state[Y][a]
= I$ and let $dir[X][a] = S, dir[Y][a] = I$. $X$ receives a store request from
the core, and hence sends an upgrade request \Req{X}{Dir}{a}{M} to the
directory. Say $X$ was a non-blocking cache, and hence receives another load
request for address $a'$ from the core before receiving an upgrade authorization
for address $a$. $X$ does not have space in the cache for $a'$, and it decides
to evict address $a$, and sends \Resp{X}{Dir}{a}{I} to the directory, changing
$state[c][a]$ to $I$.  Meanwhile, the directory receives the request
\Req{X}{Dir}{a}{M} and since $dir[Y][a] \le I$, it responds to $X$ with an
upgrade authorization \Resp{Dir}{X}{a}{M} (along with the data). Later, the
directory receives the notification \Resp{X}{Dir}{a}{I} from $X$, and downgrades
$dir[X][a]$ to $I$. Meanwhile, $X$ receives the upgrade authorization
\Resp{Dir}{X}{a}{M} and changes $state[X][a]$ to $M$. Now,
\FirstInv.\ref{conservative} is broken, since $state[X][a] = M$ and $dir[X][a] =
I$. Let $Y$ get a load request from its core, and hence sends a
\Req{Y}{Dir}{a}{S} to the directory. The directory on receiving the request,
sends an upgrade authorization to $Y$ because $dir[X][a] = I$. This violates the
single-writer property and hence coherence.

>>>>

\section{A two-level message-passing directory-based MSI protocol}
\label{sec:msi}

\newcommand{\fourAngle}[6]{\text{$#1\langle#2\rightarrow#3,#4,#5\rightarrow#6\rangle$}}
\newcommand{\threeAngle}[5]{\text{$#1\langle#2\rightarrow#3,#4,#5\rangle$}}

%\newcommand{\Req}[5]{\fourAngle{Req}{#1}{#2}{#3}{#4}{#5}}
\newcommand{\Req}[4]{\threeAngle{Req}{#1}{#2}{#3}{#4}}
\newcommand{\Resp}[4]{\threeAngle{Resp}{#1}{#2}{#3}{#4}}

\newcommand{\msg}[5]{\text{$#1 \langle #2 \rightarrow #3, #4, #5 \rangle$}}
%\newcommand{\Req}[5]{\msg{Req}{#1}{#2}{#3}{#4}{#5}}
%\newcommand{\Resp}[5]{\msg{Resp}{#1}{#2}{#3}{#4}{#5}}
\newcommand{\Data}[4]{\msg{Data}{#1}{#2}{#3}{#4}}
\newcommand{\Mesg}[4]{\msg{Mesg}{#1}{#2}{#3}{#4}}
\newcommand{\DirInv}{\emph{DirInv}}
%\input{Atomic}

Let us walk through our framework using the example of an MSI protocol.  We will
start with a simple system for the sake of explanation.  The system has two
levels of memory hierarchy -- private L1 caches for each core and a shared
memory. The L1 caches do not have direct access to the memory, and the memory
does not have a direct access to the actual states in the  L1 caches, or L1's
data. Instead, the main memory has direct access to a directory associated with
it which keeps a version of the L1 cache states for each address. The L1 caches
do not have direct access to this directory. The directory's version of the L1
cache states may differ from the actual L1 cache states. We use the terms memory
and directory interchangeably.

\begin{figure}
\begin{tabularx}{\linewidth}{|cX|}
\hline
$state[c][a]$ & coherence state (from set $\{M, S, I\}$) of cache $c$ for address $a$\\
$data[c][a]$ & data in cache $c$ for address $a$\\
\hline \end{tabularx}
\caption{Values stored by an L1 cache $c$ for an address $a$}
\label{table:lineinfo}
\end{figure}

\begin{wrapfigure}{r}{.4\linewidth}
\centering
\begin{tabular}{|c|ccc|}
\hline
& $M$ & $S$ & $I$\\
\hline
$M$ & $=$ & $>$ & $>$\\
$S$ & $<$ & $=$ & $>$\\
$I$ & $<$ & $<$ & $=$\\
\hline
\end{tabular}
\caption{$<$ and other relations for MSI states}
\label{msi<}
\end{wrapfigure}

Each L1 cache stores the information given in Figure \ref{table:lineinfo}. The
coherence state for an address denotes the permission that a cache has for the
address: $M = \{R, W\}, S = \{R\}, I = \{\}$. We can define a $<$ relation for
coherence state values based on the permissions the states represent. This is
shown in Figure \ref{msi<}. State $I$ is akin to the cache not having the data
corresponding to $a$, since it can neither read or write the $a$'s data. Figure
\ref{table:dirinfo} shows the information stored in the memory and directory.

@@

The L1 caches and the directory communicate with each using via messages. Each
L1 cache communicates with the memory directly via 4 buffers -- two incoming
buffers from the memory, one each for requests and responses from the memory,
and two outgoing buffers into the memory, one each for requests and responses to
the memory. All these buffers converge into or diverge from 4 buffers of the
memory -- two incoming buffers, one each for requests and responses from the
children and two outgoing buffers, one each for requests and responses to the
children (Figure \ref{fig:setup}). All these buffers have the FIFO property. Not
only that, we will also assume that requests sent from a source will never reach
its destination till an earlier response sent by the same source has been
handled (and hence removed) by the destination. We will show in Section
\ref{sec:fifo} about how these network requirements can be realistically
implemented.

\subsection{No replacement evictions}

We will first design an MSI protocol in which the L1 caches have enough
capacity (or rather associativity) so that no line gets evicted due to
replacement. Lines may however get evicted to enforce the following invariant
which is needed for maintaining cache coherence.

\begin{inv}
Single-writer: When a cache has permission to modify data for an address ($M$
state), no other cache has read or write permission for that address. In other
words, the other caches must be in $I$ state.
\label{singleWriter}
\end{inv}

\begin{wrapfigure}{r}{.46\linewidth}
\begin{subfigure}{.27\linewidth}
\centering
\begin{tabular}{|ccc|}
\hline
$M$ & $\rightarrow$ & $I$\\
$S$ & $\rightarrow$ & $S$\\
$I$ & $\rightarrow$ & $I$\\
\hline
\end{tabular}
\caption*{$toCompatible$}
\label{toCompat}
\end{subfigure}
~~~~~~~~~~~~~~
\begin{subfigure}{.21\linewidth}
\centering
\begin{tabular}{|c|c|}
\hline
$M$ & $\checkmark$\\
$S$ & $\times$\\
$I$ & $\times$\\
\hline
\end{tabular}
\caption*{$isModified$}
\label{isModified}
\end{subfigure}
\caption{$toCompatible$ and $isModified$ mappings}
\label{fig:msimap}
\end{wrapfigure}

Because of the Single-writer invariant, it makes sense to talk about
compatibility of the states of two caches for an address. Two cache states $x$
and $Y$ are \emph{compatible} if two caches having the same address in states
$x$ and $y$ do not violate the Single-writer invariant. We define a mapping
$toCompatible(x)$ which gives the highest state which is compatible with a given
state $x$ (Figure \ref{fig:msimap}).

In addition to Single-writer Invariant, maintaining cache coherence also
requires the following invariant:

\begin{inv}
Read-from-last-writer: When a cache reads data for an address, it reads the last
modified version of the data for the address. If the data for an address has not
been modified, the cache reads the data from memory.
\end{inv}

Function $isModified(x)$ determines if data for an address state $x$ can
potentially be modified or not (Figure \ref{fig:msimap}).

If a cache $c$ receives a request for an address $a$ from the processor, it can
service the request straightaway if it has enough permissions for address $a$.
Otherwise, it has to get to a state for $a$, say $y$, which has enough
permissions. In other words, it has to \emph{upgrade} its state to $y$. In order
to maintain the Single-writer Invariant \ref{singleWriter}, cache $c$ has to
ensure that the states of other caches are \emph{compatible} with $y$. But the
caveat is that $c$ can access only its own state and data, as opposed to that of
other caches (or the directory).

Cache $c$ has to instead communicate with the directory to ensure that the
states of other caches are compatible. $c$ must send a message
\Req{c}{Dir}{a}{y} requesting the directory to tell if it can upgrade its state
for address $a$ to $y$. We call this an \emph{upgrade request}. Let us now take
for granted that such a message can be sent (and see how this in ensured later).

\begin{inv}
A cache will eventually be able to send an upgrade request.
\label{csendreq}
\end{inv}

We will also assume that that a cache can upgrade its state to $y$ only after it
receives a message from the directory \Resp{Dir}{c}{a}{y} permitting it to
upgrade its state for address $a$ to $y$. We call such a message as an
\emph{upgrade response}.

\begin{inv}
Cache $c$ can upgrade $state[c][a]$ from $x$ to $y$ where $y > x$ only after it
receives an upgrade response \Resp{Dir}{c}{a}{y} from the directory.
\label{cwaitresp}
\end{inv}

We introduce two other kinds of messages: \emph{downgrade request}
\Req{Dir}{c'}{a}{y} sent from the directory to a cache $c'$ asking $c'$ to
downgrade its state for address $a$ to $y$, and \emph{downgrade response}
\Resp{c'}{Dir}{a}{y} sent from $c'$ to the directory informing the directory
that $c'$ has changed its state $state[c'][a]$ from $x$ to $y$, where $x > y$.

How can the behavior of the directory be designed (in terms of sending and
receiving messages) to ensure that cache $c$ will get an upgrade response
eventually. Let's say the invariants given in Figure \ref{DirInv} are
somehow guaranteed. With these invariants, let us design the behavior of the
directory.

\floatstyle{boxed}
\restylefloat{figure}
\begin{figure}
\begin{inv}
Just before receiving an upgrade request \Req{c}{Dir}{a}{y} from $c$ $dir[c][a]
< y$.
\label{drecvreq}
\end{inv}
\begin{inv}
The directory will eventually be able to send any downgrade request to any cache
$c$.
\label{dsendreq}
\end{inv}
\begin{inv}
The directory can change $dir[c][a]$ from $x$ to $y$ where $y < x$, \ie
downgrade its directory state only after receiving a downgrade response
\Resp{c}{Dir}{a}{y} from $c$.
\label{dwaitresp}
\end{inv}
\begin{inv}
If the directory has sent a downgrade request \Req{Dir}{c}{a}{y} to cache $c$,
it will eventually get a downgrade response \Resp{Dir}{c}{a}{z} from cache $c$
such that $z \le y$.
\label{drecvresp}
\end{inv}
\begin{inv}
Just before receiving a downgrade response from $c$, if $isModified(dir[c][a])$
is true, then $c$ would have sent the data along with the downgrade response.
\label{drecvdata}
\end{inv}
\begin{inv}
The directory will eventually be able to send any upgrade response to any cache
$c$. It can also send the data to cache $c$ along with the upgrade response.
\label{dsendresp}
\end{inv}
\begin{inv}
Conservative directory:
\begin{spacing}{.1}
\begin{equation*}
\forall c, a,\; dir[c][a] \ge state[c][a]
\end{equation*}
\end{spacing}
\label{conservative}
\end{inv}
\caption{Directory invariants}
\label{DirInv}
\end{figure}
\floatstyle{plain}
\restylefloat{figure}

%The directory ensures that the states of any two caches for an address are
%compatible by the following behavior.
%When the directory receives an upgrade request
%\Req{c}{Dir}{a}{x}, Invariant \ref{drecvreq} guarantees that $dir[c][a] < x$ and
%because of Invariant \ref{conservative}, $cache[c][a] le dir[c][a]$, resulting
%in $cache[c][a] < x$. Unless the directory sends an upgrade response to $c$
%permitting it to upgrade to $x$, $c$ will remain in state $cache[c][a]$ forever
On receiving an upgrade request \Req{c}{Dir}{a}{x}, the directory sends a
downgrade request \Req{Dir}{c'}{a}{toCompatible(x)} to every cache $c'$
for which $dir[c'][a] > toCompatible(x)$. Such a request can be sent eventually
because of Invariant \ref{dsendreq}. The incoming upgrade request is not
removed and no further upgrades requests are processed till an appropriate
upgrade response is sent back from the directory to $c$.

Because of Invariant \ref{drecvresp}, the directory will eventually receive
downgrade responses \Resp{c'}{Dir}{a}{z} from each of the requested caches $c'$
such that $z \le toCompatible(x)$ (since the directory made a downgrade request
to go to $toCompatible(x)$). Just before receiving the downgrade response, if
$isModified(dir[c'][a])$ is true (\ie $dir[c'][a] = M$), then by Invariant
\ref{drecvdata}, the directory would also receive data from $c'$ and the memory
for address $a$ is updated with the received data.

Once all the downgrade responses have been received, since no upgrade responses
have been sent to any cache meanwhile, we have $\forall c' \neq c, dir[c'][a] \le
toCompatible(x)$. Because of Invariant \ref{conservative}, it is guaranteed that
$\forall c' \neq, cache[c'][a] \le toCompatible(x)$. The directory now changes
$dir[c][a]$ to $x$ and sends an upgrade response \Resp{Dir}{c}{a}{x} to $c$. If
$dir[c][a] = I$ right before the upgrade response is sent, then it implies that
$cache[c][a] = I$. So the data from the memory for address $a$ is also sent
along with the upgrade response. Note that such an upgrade response, along with
the data, can always be sent eventually by the directory because of Invariant
\ref{dsendresp}. Finally, the upgrade response from $c$ is removed by the
directory.

The above behavior for the directory ensures that the states of any two caches
for every address are compatible, as can seen by induction.  This property is
true initially because every cache is in state $I$ for every address. If this
property is true before a request is started received, it will be true by the
time the request is removed and an upgrade response is sent. This is because an
upgrade response is sent only when every cache other than the one requesting is
at a state $\le toCompatible(x)$, where $x$ is the state the requesting cache
wants to go to, and $toCompatible(x)$ is the highest state that a cache can be
in to be compatible with another cache in state $x$ for an address.

Whenever the directory receives a downgrade response from a cache $c'$, just
before receiving the response if $isModified(dir[c'][a])$ is true, \ie
$dir[c'][a] = M$, $c'$ must have been the only cache with permission to modify
address $a$. This is because the directory ensures that the states of any two
caches for address $a$ is compatible at all times (as shown above). Thus, to
ensure the Read-from-last-writer Invariant, the potentially modified data from
$c'$ is written back from into the memory on receiving the downgrade response
from $c'$ as seen in our design.

The invariants \ref{drecvreq} to \ref{conservative} in Figure \ref{DirInv}
essentially makes it straight forward to design the behavior of the directory.
For the sake of argument, let us assume Invariant \ref{drecvreq} not be
guaranteed. Consider a scenario in which the cache $c$ which has sent a request
\Req{c}{Dir}{a}{M}. Before receiving an upgrade response for this request, $c$
sends another upgrade request \Req{c}{Dir}{a}{S}. Let us assume that the
directory processed the requests in the order in which they were sent. Once the
directory has finished processing the first request, it changes its directory
state $dir[c][a]$ to $M$. When it receives the second request, it is not easy to
decide if the directory has to drop the request or to send another response
authorizing the cache to change to $S$. But if Invariant \ref{drecvreq} is
guaranteed, it's guaranteed that such a scenario will never arise.

The difficulty in message passing protocols comes from the fact that it is
difficult to guarantee the above invariants, even for a 2-level memory hierarchy
discussed above as shown in the following scenarios.

%\begin{scen}
%\emph{Inability to send messages because of network congestion:}
%The ability of a cache or the directory to send messages (Invariants
%\ref{dsendreq}, \ref{dsendresp} and \ref{csendreq}) depends upon the
%availability of free buffers in the network; if the network remains congested
%all the time, then it would not be possible to send the messages, and the system
%would deadlock.
%\end{scen}
%
%\begin{scen}
%\emph{Multiple upgrade requests from the cache for the same address:}
%If a cache is allowed to make multiple requests to upgrade the state of an
%address (say to $M$ and then to $S$) before receiving the upgrade responses,
%then it might violate Invariant \ref{drecvreq} if the requests were processed by
%the directory in the order they were received. This makes the decision on
%whether the directory drops the second request or not hard to reason about, as
%shown above.
%\end{scen}

\begin{scen}
\emph{On sending an upgrade request, the cache stops processing downgrade
requests it gets an upgrade response:}
Consider a scenario with two caches $c$ and $c'$, with both the caches in state
$S$ for address $a$. Both the caches get a store request from the processor and
send \Req{c}{Dir}{a}{M} and \Req{c'}{Dir}{a}{M}, respectively, to the directory.
Let's say \Req{c'}{Dir}{a}{M} arrives at the directory first. The directory
sends a downgrade request to cache $c$, but since $c$ does not process any
downgrade requests till it gets its upgrade response, it will never send back a
downgrade response to the directory, violating Invariant \ref{drecvresp},
resulting in a deadlock. Note that this deadlock would occur even if a cache is
waiting for the upgrade response, and stops processing downgrade requests for
any address, even if the address corresponding to the downgrade request is not
the same as the one corresponding to the upgrade response.
\end{scen}

\begin{scen}
\emph{At the directory, downgrade responses from children are blocked by upgrade
requests:}
Consider a scenario where two caches $c$ and $c'$ are both in state $S$ for
address $a$. $c'$ gets a store request from the processor and sends an upgrade
request \Req{c'}{Dir}{a}{M} to the directory. The directory receives that
request, and sends a downgrade request \Req{c}{Dir}{a}{I} to $c$. $c$ sends a
downgrade authorization \Resp{c}{Dir}{a}{I} to the directory on receiving the
downgrade request. But the directory can not receive the authorization because
it is blocked by other requests (including \Req{c'}{Dir}{a}{M}). This violates
Invariant \ref{drecvresp} and hence creates a deadlock.
\end{scen}

These scenarios show that if the invariants of Figure \ref{DirInv} are violated,
then it either breaks cache coherence (by creating deadlocks) or it makes
it hard to reason about the behavior of the directory.

Before we show how the invariants of Figure \ref{DirInv} can be guaranteed, let
us consider a slightly more complicated system where caches can evict lines for
replacement.

\subsection{Replacement evictions are allowed}
If a cache receives a request from the core for an address $a$ and has no space
to accommodate address $a$, it evicts another address $a'$ to make room for $a$.

In the previous scenario, the caches didn't have to evict for replacement; a
cache had to send downgrade responses to the directory only on receiving a
corresponding request. But in the new scenario, a cache can send a downgrade
notification voluntarily to the directory on evicting an address. Such a
notification is useful in keeping the directory's version of the cache's state
up to date. We will restrict the family of directory-based protocols that that
we discuss to those in which the caches send such notifications.

Since the directory can not distinguish between the downgrade notification sent
voluntarily and a downgrade response sent because of a downgrade request from
the directory, we will use the same message to denote both.

We will first discuss an example of the behavior of the L1 cache that would
violate Invariant \ref{conservative}.

\begin{scen}
\emph{Cache sends voluntary downgrade responses indiscriminately:}
Consider a system with only two caches $c$ and $c'$ (Figure \ref{fig:cross})
such that $state[c][a] = dir[c][a] = S$, and $state[c'][a] = dir[c'][a] = I$.
Let $c$ be a non-blocking cache, so it can process requests from the core even
if there are pending misses. $c$ receives a store request for address $a$ from
the core and sends \Req{c}{Dir}{a}{M} to the directory.  Before it gets back an
upgrade response from the directory, it receives another load request for
address $a'$ from the core. Suppose the line for $a'$ is not present in $c$ and
$c$ has no space for $a'$, it decides to evict address $a$ to make room for
$a'$. It voluntarily sends a downgrade response \Resp{c}{Dir}{a}{I} to the
directory. Meanwhile, the directory receives the upgrade request
\Req{c}{Dir}{a}{M}, and since the directory state of the only other cache is $I$
for address $a$, the directory sends back an upgrade response. The directory
then receives the downgrade response \Resp{c}{Dir}{a}{I} and changes $dir[c][a]$
to $I$. $c$ finally receives the upgrade response \Resp{Dir}{c}{a}{M} and
upgrades $state[c][a]$ to $M$. This breaks Invariant \ref{conservative}, since
$state[c][a] = M$ and $dir[c][a] = I$. If the directory gets an upgrade to $M$
request for address $a$ from $c'$, it will assume that $c$ is in state $I$ for
address $a$ and send an upgrade to $M$ response to $c'$. This will violate the
Single-writer invariant, thereby violating coherence.
\end{scen}
